\documentclass[12pt]{report}

%---------Packages----------%

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage[normalem]{ulem}
\usepackage{lmodern, csquotes, graphicx, lipsum, booktabs, amsmath, setspace, rotating}

%---------TitleSec----------%

\usepackage[explicit]{titlesec}
\titleformat{\chapter}[display]{\bfseries\centering}{\Huge Chapter \thechapter:}{1em}{\Huge #1}

%---------FancyHdr----------%

%\usepackage{fancyhdr}
%\pagestyle{fancy}
%\fancyhead{}
%\fancyhead[L]{\nouppercase{\leftmark}}
%\fancyhead[R]{Sound of the Underground}
%\fancyfoot{}
%\fancyfoot[L]{Jacob P. Cupps}
%\fancyfoot[R]{\thepage}
%\renewcommand{\headrulewidth}{0.4pt}
%\renewcommand{\footrulewidth}{0.4pt}
%\renewcommand{\chaptermark}[1]{\markboth{#1}{}}

%---------BibLaTeX----------%

\usepackage[notes, backend=biber]{biblatex-chicago}
\bibliography{references.bib}

%---------Epigraph----------%

\usepackage{epigraph}
\AtBeginDocument{\renewcommand {\epigraphflush}{center}}
\setlength \epigraphwidth {0.6\linewidth}
\setlength\epigraphrule{0pt}

%---------Verse------------%

\usepackage{verse}
\renewcommand*{\theHpoemline}{\arabic{verse@envctr}.\arabic{poemline}}

%---Hyperref, after Verse--%

\usepackage[hidelinks]{hyperref}

%----------DOCUMENT---------%

\begin{document}
\pagenumbering{roman}
\thispagestyle{empty}
\input{chapters/titlepage}
\frenchspacing %\doublespacing

%-------Front Matter--------%

\setcounter{page}{2}
  %----_TOC, Lists-----%
    \tableofcontents
    \listoffigures
    \addcontentsline{toc}{chapter}{List of Figures}
    \listoftables
    \addcontentsline{toc}{chapter}{List of Tables}

  %--Acknowledgements--%
    \chapter*{Acknowledgements}
    \addcontentsline{toc}{chapter}{Acknowledgements}

This project is indebted to an always-growing number of people who influence, encourage, 
and inspire me. First, I would like to thank my advisor for this project, Paul Steinbeck, 
as well as my second readers, Lauren Eldridge Stewart and Christopher Stark. The classes 
I've taken as well as the time I've spent in discussion with each of you have helped shape
this project in distinct and insightful ways. My thanks also goes out to other academic 
mentors I've connected with at Washington University in St. Louis. I extend my thanks to 
Paula Harper and Alexander Stefaniak, both of whom read and provided feedback on papers 
and abstracts that informed my research early on in its development.
 
I adamantly believe this project would not be what it is without the camaraderie of the 
graduate students in Washington University's department of music. I am thankful to all my 
colleagues for the many hours of discussion and collaboration inside and outside of class,
in addition to their commitment toward making our department an inclusive and amenable space.
My special thanks go out to my Eden Attar and Varun Chandrasekhar; I appreciate both of you
for the time spent writing with me, listening to the music I studied, and giving me space to
workshop my ideas.

My friends and family have created an integral community of support around me throughout my 
completion of this project and my time in academia in general. There are more folks to name 
than just the ones here, but a few who have been indispensable to me are my partner Haley 
Jo Mack and my friend Emory Parker; you both have been there for me as readers, confidants, 
and\textemdash more than anything\textemdash a support system. I am similarly thankful to my 
brother and father, Justin and Brian Cupps: you two keep me grounded and have helped me
immensely throughout my time in this master's program. I would like to dedicate this work to
the four of you, for, truly, I am able to do what I do because of how you all have shaped me.



\vspace{0.2cm}
\hfill{Jacob P. Cupps}

\noindent \textit{Washington University in St. Louis}

\noindent \textit{May 2022}

  %-----Abstract-----%
    \chapter*{Abstract}
    
    \begin{center}
        \large ``The Sound of the Underground'': \\
        \large Distinguishing Alternative Identity in Rap's Contemporary Moment  \\
        \normalsize by \\
        Jacob P. Cupps \\
        Master of Arts in Music Theory \\
        
        Washington University in St. Louis, 2022 \\
    \end{center}
    
    \addcontentsline{toc}{chapter}{Abstract}

    \normalsize
    This thesis focuses on transcribing and analyzing a repertory of underground rap songs, examining
    how the quality of undergroundness manifests within beats and flows. I define underground as a
    subgenre of hip-hop that is unified by a culture of meditative listening, and I opt for it over
    related terms because it is style agnostic. Therefore, I am able to investigate how a small set
    of techniques manifest across a wide variety of approaches to making underground hip-hop. I argue
    that these techniques are a method of sounding the underground, communicating meaning to the 
    listener at the level of the song.

    After an introductory Chapter 1, the subsequent chapters examine techniques used in underground
    production and underground emceeing respectively. Chapter 2 argues that producers employ techniques
    I term resampling/recomposing, choking, glitching, and slipping to create beats that use non-standard
    song forms and prioritize an aesthetic of internal variety and contrast; this heterogeneous sound
    promotes producers to co-authors in the construction of musical identity. Chapter 3 posits that emcees
    use techniques I term pivot rhyme, closing fragmentation, mimesis, and processing to engage listener's
    expectations and challenge the perception that flow is the most hierarchically significant element in a
    rap song. This demonstrates an emcee's reciprocity in the construction of their own identity, inviting
    listeners to deeply engage with how the music informs the resulting musical persona.


%---------------------------%
\chapter{Locating the Underground in Styles, Histories, and Practices}
\pagenumbering{arabic}
\input{chapters/chp-01}

\chapter{Beat Construction and Methods of Alteration in the Underground}
\input{chapters/chp-02}

\chapter{Structural and Performance Tools for the Underground Emcee}
\input{chapters/chp-03}

%---------------------------%

\singlespacing
\printbibliography
\addcontentsline{toc}{chapter}{Bibliography}
\nocite{*}

%---------------------------%
\chapter*{Appendix}
\addcontentsline{toc}{chapter}{Appendix}
\input{chapters/app-01}


\end{document}
